Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-08-16 paper/main.tex
[DATALAD RUNCMD] run codespell throughout but ignore fail
=== Do not change lines below ===
{
 "chain": [],
 "cmd": "codespell -w || :",
 "exit": 0,
 "extra_inputs": [],
 "inputs": [],
 "outputs": [],
 "pwd": "."
}
^^^ Do not change lines above ^^^

%%%%%%%% JEREMY VERSION %%%%%%%%%
%Installing new packages in a notebook using standard approaches (e.g., system commmands) affect the
%runtime environment. This could lead to undesired behaviors. For example, running a notebook that
%installs new packages in the user’s primary system environment might alter their main system installation
%and/or containerized environment in unexpected ways (e.g., changing package versions, causing conflicts
%with other packages, etc.). To protect against undesired changes to the runtime environment, Davos
%incorporates its own virtual environment for managing packages it installs. When Davos is imported,
%a new virtual environment (folder) is created automatically. The folder’s name may be customized to
%support multi-notebook projects. Any smuggled packages that were not available in the notebook’s
%runtime environment are installed to the current project folder. The runtime environment remains unaffected
%by Davos’s behavior (see Sec. 2.2.3).

@@

In lines 26--29, we demonstrate another aspect of Davos's
functionality that supports more advanced installation scenarios.  The
\texttt{ipywidgets}~\cite{FredEtal15} package provides an API for
creating various JavaScript widgets with Python code, and the \texttt{widgetsnbextension} package provides
the machinery needed by the notebook frontend to display them.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example6}
\end{center}
A complication is that \texttt{ipywidgets} must be installed in the
same environment as the IPython kernel, whereas
\texttt{widgetsnbextension} must be installed in the environment that
houses the Jupyter notebook server. In many basic setups, these two
environments are the same.  However, a common ``advanced'' approach
entails running the notebook server from a base environment, with
additional environments each providing their own separate,
interchangeable IPython kernels.  To accomodate this multi-environment
scenario, on lines 26 and 28, we use the \texttt{pip\_executable} option to control which environments each
package should be installed to.  Once these two packages are installed
and imported, line 31 smuggles \texttt{tqdm}~\cite{daCoEtal22}, which
display progress bars to provide status updates for running code. In
Jupyter notebooks, the \texttt{tqdm.notebook} module can be imported
to enable more aesthetically pleasing progress bars that are displayed via
\texttt{ipywidgets}, if that package is installed and
importable. Therefore, to take advantage of this feature, it was
important to \texttt{smuggle} \texttt{tqdm} after ensuring the
\texttt{ipywidgets} package was available.

@@

Although existing notebooks \textit{can} incorporate system calls that install
project requirements, handling project requirements in the general case is
non-trivial (e.g., see Fig.~\ref{fig:flow-chart}). Further, Davos
incorporates its own virtual environment system that isolates
notebook-installed packages from the runtime environment
(Sec.~\ref{subsec:projects}). In many setups this feature can eliminate the
need to set up a seperate virtual environment or container (e.g., in
conjunction with a \texttt{requirements.txt}, \texttt{project.toml}, or
\texttt{environment.yml} file specifying the project's dependencies).

>>>>

%%%%%%%% JEREMY VERSION %%%%%%%%%
%Installing new packages in a notebook using standard approaches (e.g., system commands) affect the
%runtime environment. This could lead to undesired behaviors. For example, running a notebook that
%installs new packages in the user’s primary system environment might alter their main system installation
%and/or containerized environment in unexpected ways (e.g., changing package versions, causing conflicts
%with other packages, etc.). To protect against undesired changes to the runtime environment, Davos
%incorporates its own virtual environment for managing packages it installs. When Davos is imported,
%a new virtual environment (folder) is created automatically. The folder’s name may be customized to
%support multi-notebook projects. Any smuggled packages that were not available in the notebook’s
%runtime environment are installed to the current project folder. The runtime environment remains unaffected
%by Davos’s behavior (see Sec. 2.2.3).

@@

In lines 26--29, we demonstrate another aspect of Davos's
functionality that supports more advanced installation scenarios.  The
\texttt{ipywidgets}~\cite{FredEtal15} package provides an API for
creating various JavaScript widgets with Python code, and the \texttt{widgetsnbextension} package provides
the machinery needed by the notebook frontend to display them.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example6}
\end{center}
A complication is that \texttt{ipywidgets} must be installed in the
same environment as the IPython kernel, whereas
\texttt{widgetsnbextension} must be installed in the environment that
houses the Jupyter notebook server. In many basic setups, these two
environments are the same.  However, a common ``advanced'' approach
entails running the notebook server from a base environment, with
additional environments each providing their own separate,
interchangeable IPython kernels.  To accommodate this multi-environment
scenario, on lines 26 and 28, we use the \texttt{pip\_executable} option to control which environments each
package should be installed to.  Once these two packages are installed
and imported, line 31 smuggles \texttt{tqdm}~\cite{daCoEtal22}, which
display progress bars to provide status updates for running code. In
Jupyter notebooks, the \texttt{tqdm.notebook} module can be imported
to enable more aesthetically pleasing progress bars that are displayed via
\texttt{ipywidgets}, if that package is installed and
importable. Therefore, to take advantage of this feature, it was
important to \texttt{smuggle} \texttt{tqdm} after ensuring the
\texttt{ipywidgets} package was available.

@@

Although existing notebooks \textit{can} incorporate system calls that install
project requirements, handling project requirements in the general case is
non-trivial (e.g., see Fig.~\ref{fig:flow-chart}). Further, Davos
incorporates its own virtual environment system that isolates
notebook-installed packages from the runtime environment
(Sec.~\ref{subsec:projects}). In many setups this feature can eliminate the
need to set up a separate virtual environment or container (e.g., in
conjunction with a \texttt{requirements.txt}, \texttt{project.toml}, or
\texttt{environment.yml} file specifying the project's dependencies).
